Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-05-26 SIR_manuscript.tex
more comments

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of test return, i.e., rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & \percap testing intensity & 1/day & - \\ \hline 
  $\theta\_w$ & Isolation efficacy (reduction of the transmission probability) for ``waiting'' individuals & - & - \\ \hline
  $\theta\_c$ & Isolation efficacy ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.339 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $w_S, w_I, w_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

@@

The low-testing case \fref{pan} reflects the case where $rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is $\approx 1.017$, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing. \bmb{how did we choose this particular value?}

@@

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at high level of \percap testing intensity $\rho$.}
We numerically evaluate the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to testing and isolation. We use the following parameters (see Table~\ref{tab:params} for a brief description):
$N_0=1 \times 10^6$, $\omega \in [1/5,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,1/5)$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.339$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The black dotted line in each panel represents the critical contour of $\Delta=1-\frac{\gamma}{\beta}$, i.e., the $\Delta$ corresponding to the threshold of $\Rnum=1$.}
\label{pan2}
\end{figure}

@@

\begin{itemize}
\item When waiting individuals do not isolate (e.g. $\theta_w =0$, top row of \fref{pan}), increasing testing delays (i.e. decreasing $\omega$) increases \Rnum, as we might expect: $\Delta$ decreases --- a larger region of the parameter space has $\Rnum>1$, and the threshold testing intensity required to control the epidemic increases --- as we move to the right within each plot in this row. However, when waiting individuals are at least partially isolated ($\theta_w>0$), we typically see the opposite effect: longer testing delays lead to \emph{less} transmission. This is because returning negative tests leads people to stop distancing; this applies both to susceptibles, and to infectious people who receive negative test results because they were sampled when susceptible (or because of imperfect test sensitivity, which we have not considered in our analysis). For small but non-zero waiting-isolation levels (e.g. $\theta_w=0.25$, second row from top), with high levels of confirmed isolation ($\theta_c \ge 0.5$), increasing testing delays initially decreases $\Delta$, then increases it as the testing delays pass a threshold value of approximately 2--5 days.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, but only for an unrealistic scenario --- very high testing intensities \emph{that begin before the epidemic starts}. It occurs when waiting people do not isolate ($\theta_w=0$), and most easily for targeted testing (\fref{pan2}(b), right three panels of top row) \bmb{can we rule out the possibility that this phenomenon could happen with random testing? In the left panel of \fref{pan2}, $\theta_w=0$/$\theta_c=0.75$ panel, there is a hint of this behaviour; it looks like the light green [second lightest] region displays this behaviour, but it's hard to see even a maximum magnification}. This phenomenon occurs --- increasing testing intensity \emph{increases} \Rnum --- because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; these people must then wait for their tests to be returned (during which time they may become infected) before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.

>>>>

\begin{table}[htp]
\centering
\begin{tabular}{|c|L{2in}cc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of test return, i.e., rate of onward flow from ``waiting'' to ``confirmed'' or ``untested'' compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & \percap testing intensity & 1/day & - \\ \hline 
  $\theta\_w$ & Isolation efficacy (reduction of the transmission probability) for ``waiting'' individuals & - & - \\ \hline
  $\theta\_c$ & Isolation efficacy for ``confirmed positive'' individuals & - & -  \\ \hline
  $\beta$ & Transmission rate & 1/day & 0.339 \\ \hline
  $\Lambda$ & Force of infection & 1/day & - \\ \hline
  $p_S$ & Probability of positive tests for $S$ ($= 1-\textrm{specificity}$) & - & 0 \\ \hline
  $p_I$ & Probability of positive tests for $I$ ($= \textrm{sensitivity}$) & - & 1 \\ \hline
  $p_R$ & Probability of positive tests for $R$ ($= 1-\textrm{specificity}$) & - & 0.5 \\ \hline
  $w_S, w_I, w_R$ & Relative testing weights & - &
  \begin{minipage}[t]{0.21\columnwidth}%
    Random: $\{1,1,1\}$ \\ Targeted: $\{0.3,1,1\}$
  \end{minipage} \\
  \hline
  \end{tabular}
\caption{\label{tab:params} Parameters of the model \eqref{model}.}
\end{table}

@@

The low-testing case \fref{pan} reflects the case where $rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
  The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is $\approx 1.017$, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing. \bmb{how did we choose this particular value?}
  \bmb{JD wonders whether we really need the ``critical'' $\Rnum=1$ contour or not; any contour will do to illustrate a threshold level of testing intensity, and the fact that $\Rnum=1$ falls in the middle of our range is a bit arbitrary (i.e. the fact that testing can control the epidemic depends on the baseline $\Rnum$ value we have chosen). (I know it was my idea in the first place, but maybe we could dispense with it?)}

@@

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.45\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at high level of \percap testing intensity $\rho$.}
We numerically evaluate the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to testing and isolation. We use the following parameters (see Table~\ref{tab:params} for a brief description):
$N_0=1 \times 10^6$, $\omega \in [1/5,2]$ 1/day, $1/\gamma= 3$ days, $\rho \in [0,1/5)$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.339$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The black dotted line in each panel represents the critical contour of $\Delta=1-\frac{\gamma}{\beta}$, i.e., the $\Delta$ corresponding to the threshold of $\Rnum=1$. \bmb{consider removing dotted line (see note elsewhere); on this and prev figure consider removing redundant y-axis label on right panel (and adjusting/squashing interpanel spacing)?}}
\label{pan2}
\end{figure}

@@

\begin{itemize}
\item When waiting individuals do not isolate (e.g. $\theta_w =0$, top row of \fref{pan}), increasing testing delays (i.e. decreasing $\omega$) increases \Rnum, as we might expect: $\Delta$ decreases --- a larger region of the parameter space has $\Rnum>1$, and the threshold testing intensity required to control the epidemic increases --- as we move to the right within each plot in this row. However, when waiting individuals are at least partially isolated ($\theta_w>0$), we typically see the opposite effect: longer testing delays lead to \emph{less} transmission. This is because returning negative tests leads people to stop distancing; this applies both to susceptibles, and to infectious people who receive negative test results because they were sampled when susceptible (or because of imperfect test sensitivity, which we have not considered in our analysis). For small but non-zero waiting-isolation levels (e.g. $\theta_w=0.25$, second row from top), with high levels of confirmed isolation ($\theta_c \ge 0.5$), increasing testing delays initially decreases $\Delta$, then increases it as the testing delays pass a threshold value of approximately 2--5 days.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, but only for an unrealistic scenario --- very high testing intensities \emph{that begin before the epidemic starts}. It occurs when waiting people do not isolate ($\theta_w=0$), and most easily for targeted testing (\fref{pan2}(b), right three panels of top row) \bmb{can we rule out the possibility that this phenomenon could happen with random testing? In the left panel of \fref{pan2}, $\theta_w=0$/$\theta_c=0.75$ panel, there is a hint of this behaviour; it looks like the light green [second lightest] region displays this behaviour, but it's hard to see even a maximum magnification}. This phenomenon occurs --- increasing testing intensity \emph{increases} \Rnum --- because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; these people must then wait for their tests to be returned (during which time they may become infected) before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.
  \bmb{I am still trying to understand why slowing testing return (decreasing $\omega$) decreases \Rnum over \textbf{most} of the parameter space considered here, which is a counterintuitive/silly result. Fast test returns help when we have high test positivity and low isolation during the waiting period (specifically, [much] lower isolation while waiting than when confirmed, but I think we're assuming that); we want to get waiting people into the ``confirmed'' compartment as quickly as possible. Can we understand/explain this result more thoroughly using that logic, and can we argue that what we are looking at is \emph{not} a realistic parameter range? Does the low baseline \Rnum we are using play into this, e.g. by reducing test positivity? Is there a simple expression for test positivity during the invasion phase?}
